No effects of COVID-19 related social media use on well-being

### 2021-11-03 manuscript.Rmd
- updates before sending to Nik

---
title             : "No effects of COVID-related social media use on well-being"
shorttitle        : "COVID-related social media use and well-being"

@@

abstract          : "In times of crisis such as the Corona pandemic, it is important that citizens stay informed about recent events, the latest political decisions, or mandatory protection measures. As a result, to attain relevant information many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-related topics might put personal well-being at risk. To this end, I analyze data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. The data were analzed using random effects cross lagged panel models, controlling for several stable and varying covariates. Results showed that COVID-related social media use did not meaningfully affect several types of well-being, including life satisfaction, positive affect, and negative effect. This pertains to both passive and active social media use, and all the prominent channels such as Facebook, WhatsApp, or YouTube. As a result, this study suggests fears that social media use during times of crisis might be detrimental for well-being can be put to rest."
keywords          : "COVID, well-being, affect, life satisfaction, social media use, news use, random effects within between model, panel study, longitudinal"

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

@@

During the COVID pandemic, 
<!-- and especially in the beginning,  -->
several events unfolded in quick succession. 
How dangerous is the virus? Is it spreading in my region? How is it transmitted, and how can I protect myself?
Because for many it was (and still is) a matter of life or death, citizens had to stay informed regarding the latest developments.
Governments around the world implemented safety measures, from wearing masks or keeping physical distance, to complete lockdowns.
In this extraordinary situation, people hence used media excessively, and especially social media were at an all time high [@statistaAverageDailyTime2021].
A new phenomenon emerged, "termed doomscrolling":
People could not stop using social media to attain COVID-related news.

Several people reported that they were glued to their screens and could not pursue other relevant activities such as working, taking a break, or even care-work.
Increasingly, it was asked whether such an increase in social media use could still be considered useful and adaptive, or whether it created an additional and new psychological danger for the users' mental health [@kleinDarklySoothingCompulsion2021].
A study with 6,233 people from Germany found that "[f]requency, duration and diversity of media exposure were positively associated with more symptoms of depression and unspecific and COVID specific anxiety" [@bendauAssociationsCOVID19Related2021].

As a result, with this study I want to build on this research and investigate the question whether COVID-related social media use during the pandemic affected the users' well-being.
To this end I analyze a large-scale panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of 24 waves and an overall sample size of `r nrow(d_wide)`, with at least 1,500 participants per wave, and it is representative of the Austrian population.
The panel study collected a large number of variables.
Because we can therefore control for many confounding third variables, both stable and varying, together with the longitudinal design this creates a unique opportunity to analyze causality.

@@

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology:
first, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); sixth, the message (e.g., content) [@meierComputerMediatedCommunicationSocial2020].
Whereas the first four levels are focused on the channel, the last two on the type of communication.
To measure social media use for consumption of specific news, I here employ both the channel and the communication perspective.
First, I will analyze how using the most prominent branded applications affect well-being, and whether this effect changes across applications.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
But because adopting only this position would be both too narrow and too general, I will secondly also investigate how different types of interaction affect well-being.
Specifically, I will differentiate between active and passive use.
I will distinguish (a) _reading_ COVID-related social media use (passive), _posting_ content regarding COVID (active), and _liking and sharing_ COVID-related posts by others (both active and passive).
Worth noting, this study is not about _general_ social media during times of COVID, but on social media use focused on COVID-related news and interactions.
For example, posting about the pandemic or retweeting COVID-related posts.

@@

In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-related anxiety symptoms" (p. 288).
Hence, this might hint at potential negative effects of social media as news use---but note that this finding comes from a between-person relation stemming from a cross-sectional data.
Hence, we don't know whether the differences in mental health and well-being are due to social media use or other third variables, such as age or education.

@@

What determines a trivial or a small effect is a difficult question (see below).
If we refer to standardized effect sizes, according to Cohen [@cohenPowerPrimer1992] small effect sizes start at _r_ = .10.
However, most meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputerMediatedCommunicationSocial2020].
As a result, I think it's most convincing to expect trivial to small effects also in the case of COVID-related social media use.

From a theoretical perspective, why could COVID-related social media use be detrimental?
Above everything, one can reasonably assume a _direct_ negative effect on well-being, especially on positive or negative effect.
Dangers, inequalities, corruption -- these were the headlines across many countries worldwide.
If one learns about such things, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news might be depressing, perhaps even changing some general perspective on life.
So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well.

@@

On the other hand, one can make the case that overuse can be beneficial _in times of a pandemic_, even if it's mainly COVID-related.
Exchanging COVID-related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is logically impossible at that time.
At times where meaningful and functional activities are explicitly or implicitly prohibited, using social media to exchange about COVID might not be the worst idea.
In fact, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand information on the current developments.
On the other hand, there is of course also much disinformation, and "bingeing" on COVID fake news might also pose risks for impaired well-being.

Together, the strongest argument seems to be that the effects of social media on well-being are, on average, generally small at best.
Because this study looks at only _one part_ of social media use---namely, COVID-related interactions---it is even more focused, and the overall potential of the effects should diminish even further.
Whether or not using social media for COVID-related aspects is detrimental during a pandemic is also not entirely clear.
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->
Therefore, I expect to find that the effects of COVID-related news use on well-being will be not be meaningful or practically relevant.

> Hypothesis: The within-person effects of all types of COVID-related social media use on all types of well-being indicators---while controlling for several stable and varying covariates such as sociodemographic variables and psychological dispositions---will be trivial.

@@

If a heavy user of COVID-related social media news stops using social media altogether, this should have a noticeable impact on their overall well-being.

What would this mean practically and how could it be operationalized?
In this study, COVID-related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, the example from above would imply that a change of five units in social media use should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably differentiate between _seven_ levels of satisfaction with health.
So we could state that a five unit change in social media use should result in a one unit change in satisfaction.
Statistically, in a regression, _b_ estimates the change in the dependent variable if the independent variable increases by one point.
Transferred to our example, we would hence expect a _b_ of 1 / 5 = .20.

@@

Thus, only if we can control for _all_ potential varying third variables, we can correctly estimate causality without any bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure to have included all varying covariates, which makes absolute statements regarding causality impossible.
However, when controlling for many varying variables, we can be much more certain that we measured causality correctly.
The aim should still be to collect as many relevant varying and nonvarying third variables as possible, while knowing that absolute certainty can rarely be reached.

@@

[Within vs. between logic]

[Time Frame / REWB logic]

@@

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, p-value) were preregistered on the Open Science Framework. 
The (anonymous) preregistration can be accessed here: https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3.
Because in this study I analyze data from an already existing large scale data set, the Austrian Corona Panel Project, all of these steps were preregistered _prior_ to accessing the data.
The preregistration was designed on the basis of the panel documentation online [@kittelAustrianCoronaPanel2020].
In some cases the analyses could not be executed as originally planned, because some properties of the variables only became apparent upon data analysis.
The most relevant deviations are reported below, and a complete list of all changes can be found online.

@@

The sample was representative of the Austrian population in terms of age, gender, region/state, municipality size, and educational level. In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years old. 
The average age was `r mean(2021 - d_wide$year_birth, na.rm = T) %>% round(0)`, `r mean(d_wide$male, na.rm = T) %>% round(., 2) * 100` percent were male, `r d_wide %>% filter(.$edu_fac %in% c("Hochschulverwandte Lehranstalt oder Kolleg","Bachelor", "Magister / Master / Diplomingenieur / Fachhochschule", "Doktor / PhD")) %>% nrow() / sum(table(d_wide$edu_fac)) %>% divide_by(100)` percent had a University degree, and `r table(d_wide$employment_fac)["Arbeitslos"] / (table(d_wide$employment_fac) %>% sum()) %>% unname() %>% divide_by(100) %>% round(., 0)` percent were currently unemployed.
<!-- The study received IRB approval from the University of Vienna. -->

Because the data were analyzed post hoc, no sample size planning on the basis of a priori power analyses was conducted.
Because the sample is very large, it is well-equipped reliably to detect also small effects, rendering exact post hoc power analysis unnecessary.
And also because such large sample easily generate significant p-values for very small effects, this study builds on a smallest effect size of interest.
We will use the interval testing approach as proposed by @dienesUsingBayesGet2014.
On the basis of our SESOI, we will define a null region.
For well-being, the null region will be between _b_ = -.30 and _b_ = .30.
If the 95% confidence interval lies completely within the null-region (e.g., _b_ = .15, [95% CI: .05, .25]), the hypothesis that the effect is only trivial is be supported.
If the effects interval and the null region overlap (e.g., _b_ = .25, [95% CI: .15, .35]), the hypothesis is not supported and the results are considered inconclusive, whereas a meaningful negative effect is rejected.
If the confidence falls completely outside the null-region (e.g., _b_ = .4, [95% CI: .35, .45]), the hypothesis is rejected and the existence of a meaningful positive effect supported.

@@

The hypothesis was analyzed using mixed effects models, namely random effect within-between models (REWB)[@bellFixedRandomEffects2019]. 
Three models were run, one for each dependent variable.
All predictors---that is, social media activities and social media channels, within and between-person factors, stably and varying covariates---were included in each of the three models. 
The factorial validity of the scales were tested with confirmatory factor analyses (CFA). 
If Mardia’s test showed that the assumption of multivariate normality was violated, the more robust Satorra-Bentler scaled and mean-adjusted test statistic (MLM) was used as estimator. 
To avoid overfitting, scales were tested on more liberal fit criteria (CFI > .90, TLI > .90, RMSEA <. .10, SRMR < .10) [@klinePrinciplesPracticeStructural2016].
Because REWB-models cannot model scales as latent variables, to increase precision factor scores exported from the CFAs were used.
For more information on the analyses, see [companion website](xxx).

@@

In what follows, I briefly list all variables that were collected.
For a complete list of all items and item characteristics, see [companion website](xxx).

@@

@@

The effects of COVID-related social media use were controlled for the following stable variables: 
(a) gender (answer options: female, male, diverse), (b) age, (c) education (10 ordinal options), (d) Austria country of birth (yes/no), (e) Austria parents' country of birth (no parent, one parent, both parents).
We controlled also for the following varying covariates:
(a) text-based media news consumption, (b) video-based media news consumption, (c) residency is Vienna, (d) household size, (e) Self-reported physical health, (f) Living space (in squaremeter), (g) access to balcony, (h) access to garden, (i) Employment status, (j) Work hours per week, (k) Home office, (l) household income, (m) outdoor activities, (o) satisfaction with democracy, (p) disposition to take risks, (q) locus of control.

@@

When looking at the variables from a descriptive perspective, we see that all well-being measures did not change substantially across the different phases of the pandemic.
COVID-related media use, however, increased during the beginning of the study and remained stable after approximately six waves.

@@

The study's hypothesis was that the effects of all types of social media use on well-being will be trivial.
Regarding the different types of _communication_, that is reading vs. sharing vs. posting, all within-person effects fell completely within the a-priori defined SESOIs.
For example, respondents who used social media more frequently than usual to read about COVID-related topics did not show a simultaneous change in life satisfaction (_b_ = `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within") %>% select(conf.high) %>% round(2)`]).
Only one effect did not include zero.
If people posted more about COVID-related aspects than they usually did, life satisfaction increased (_b_ = `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within") %>% select(conf.high) %>% round(2)`]).
However, because the effect did not fall outside our null region, it's likely not large enough to be considered meaningful.
As a result, the hypothesis was supported for all types of COVID-related social media communication.

Regarding between-person relations, about which no hypotheses were formulated, only two effects didn't include zero.
First, respondents who across all waves used social media more frequently than others to read about COVID reported higher levels of positive affect than others (_b_ = `r dat_fig_results_activity %>% filter(dv == "Positive affect" & iv == "Reading" & type == "between") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Positive affect" & iv == "Reading" & type == "between") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Positive affect" & iv == "Reading" & type == "between") %>% select(conf.high) %>% round(2)`]).
However, note that this effect still fell completely within the null-region.
Hence, although positive the effect was considered too small to matter practically.
Second, respondents who across all waves used social media more frequently than others to read about COVID reported lower levels of negative affect than others (_b_ = `r dat_fig_results_activity %>% filter(dv == "Negative affect" & iv == "Reading" & type == "between") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Negative affect" & iv == "Reading" & type == "between") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Negative affect" & iv == "Reading" & type == "between") %>% select(conf.high) %>% round(2)`]).
The effect was partially outside the null region, hence the effect might be large enough to be practically relevant, and the effect that it's trivial cannot be rejected.

@@

Using a panel study with a representative sample of the Austrian population that consistent of 24 waves, this study analyzed the effects of COVID-related social media use on well-being.
A random effect model that separated between person relations from within-person effects and that controlled for several third variables showed that within-person effects were trivial.
People who used social media more than usual to learn about COVID did not show changes in their well-being.
As a result, the results imply that COVID-related social media use is irrelevant for people's well-being.
Other factors among the third variables that were measured revealed much larger effects, implying that the relevance of specific types of social media use for well-being is limited.
Popular fears that "doomscrolling" or overusing social media during times of crises might not be justified.

>>>>

---
title             : "No effects of COVID-19 related social media use on well-being"
shorttitle        : "COVID-19 related social media use and well-being"

@@

abstract          : "In times of crisis such as the Corona pandemic, it is important that citizens stay informed about recent events, the latest political decisions, or mandatory protection measures. As a result, to attain relevant information many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-19 related topics might put personal well-being at risk. To this end, I analyze data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. The data were analzed using random effects cross lagged panel models, controlling for several stable and varying covariates. Results showed that COVID-19 related social media use did not meaningfully affect several types of well-being, including life satisfaction, positive affect, and negative effect. This pertains to both passive and active social media use, and all the prominent channels such as Facebook, WhatsApp, or YouTube. As a result, this study suggests fears that social media use during times of crisis might be detrimental for well-being can be put to rest."
keywords          : "COVID, well-being, affect, life satisfaction, social media use, news use, random effects within between model, panel study, longitudinal"

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
draft             : no

@@

During the COVID-19 pandemic, 
<!-- and especially in the beginning,  -->
several events unfolded in quick succession. 
How dangerous is the virus? Is it spreading in my region? How is it transmitted, and how can I protect myself?
Because for many it was (and still is) a matter of life or death, citizens had to stay informed regarding the latest developments.
Governments around the world implemented safety measures, from wearing masks or keeping physical distance, to complete lockdowns.
In this extraordinary situation, people hence used media excessively, and especially social media were at an all time high [@statistaAverageDailyTime2021].
A new phenomenon "termed doomscrolling" emerged:
People could not stop using social media to attain COVID-19 related news.

Several people reported that they were glued to their screens and could not pursue other relevant activities such as working, taking a break, or even care-work.
Increasingly, it was asked whether such an increase in social media use could still be considered useful and adaptive, or whether it created an additional and new psychological danger for the users' mental health [@kleinDarklySoothingCompulsion2021].
A study with 6,233 people from Germany found that "[f]requency, duration and diversity of media exposure were positively associated with more symptoms of depression and unspecific and COVID-19 specific anxiety" [@bendauAssociationsCOVID19Related2021].

As a result, with this study I want to build on this research and investigate the question whether COVID-19 related social media use during the pandemic affected the users' well-being.
To this end I analyze a large-scale panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of 24 waves and an overall sample size of `r nrow(d_wide)`, with at least 1,500 participants per wave, and it is representative of the Austrian population.
The panel study collected a large number of variables.
Because we can therefore control for many confounding third variables, both stable and varying, together with the longitudinal design this creates a unique opportunity to analyze causality.

@@

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology:
first, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); sixth, the message (e.g., content) [@meierComputerMediatedCommunicationSocial2020].
Whereas the first four levels are focused on the channel, the last two on the type of communication.
To measure social media use for consumption of specific news, I here employ both the channel and the communication perspective.
First, I will analyze how using the most prominent branded applications affect well-being, and whether this effect changes across applications.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
But because adopting only this position would be both too narrow and too general, I will secondly also investigate how different types of interaction affect well-being.
Specifically, I will differentiate between active and passive use.
I will distinguish (a) _reading_ COVID-19 related social media use (passive), _posting_ content regarding COVID (active), and _liking and sharing_ COVID-19 related posts by others (both active and passive).
Worth noting, this study is not about _general_ social media during times of COVID, but on social media use focused on COVID-19 related news and interactions.
For example, posting about the pandemic or retweeting COVID-19 related posts.

@@

In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
Hence, this might hint at potential negative effects of social media as news use---but note that this finding comes from a between-person relation stemming from a cross-sectional data.
Hence, we don't know whether the differences in mental health and well-being are due to social media use or other third variables, such as age or education.

@@

What determines a trivial or a small effect is a difficult question (see below).
If we refer to standardized effect sizes, according to Cohen [@cohenPowerPrimer1992] small effect sizes start at _r_ = .10.
However, most meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputerMediatedCommunicationSocial2020; @fergusonThisMetaanalysisScreen2021].
As a result, I think it's most convincing to expect trivial to small effects also in the case of COVID-19 related social media use.

From a theoretical perspective, why could COVID-19 related social media use be detrimental?
Above everything, one can reasonably assume a _direct_ negative effect on well-being, especially on positive or negative effect.
Dangers, inequalities, corruption -- these were the headlines across many countries worldwide.
If one learns about such things, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news might be depressing, perhaps even changing some general perspective on life.
So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well.

@@

On the other hand, one can make the case that overuse can be beneficial _in times of a pandemic_, even if it's mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is logically impossible at that time.
At times where meaningful and functional activities are explicitly or implicitly prohibited, using social media to exchange about COVID-19 might not be the worst idea.
In fact, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand information on the current developments.
On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being.

Together, the strongest argument seems to be that the effects of social media on well-being are, on average, generally small at best.
Because this study looks at only _one part_ of social media use---namely, COVID-19 related interactions---it is even more focused, and the overall potential of the effects should diminish even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->
Therefore, I expect to find that the effects of COVID-19 related news use on well-being will be not be meaningful or practically relevant.

> Hypothesis: The within-person effects of all types of COVID-19 related social media use on all types of well-being indicators---while controlling for several stable and varying covariates such as sociodemographic variables and psychological dispositions---will be trivial.

@@

If a heavy user of COVID-19 related social media news stops using social media altogether, this should have a noticeable impact on their overall well-being.

What would this mean practically and how could it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = _never_ to 5 = _several times a day_. 
Thus, the example from above would imply that a change of five units in social media use should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to @normanInterpretationChangesHealthrelated2003, people can reliably differentiate between _seven_ levels of satisfaction with health.
So we could state that a five unit change in social media use should result in a one unit change in satisfaction.
Statistically, in a regression, _b_ estimates the change in the dependent variable if the independent variable increases by one point.
Transferred to our example, we would hence expect a _b_ of 1 / 5 = .20.

@@

Thus, only if we can control for _all_ potential varying third variables, we can correctly estimate causality without any bias [@rohrerThinkingClearlyCorrelations2018].
Obviously, we can never be entirely sure to have included all varying covariates, which makes absolute statements regarding causality impossible.
However, when controlling for many varying variables, we can be much more certain that we measured causality correctly.
The aim should still be to collect as many relevant varying and nonvarying third variables as possible, while knowing that absolute certainty cannot be reached.

@@

Next to including covariates, it is now increasingly understood that causal effects need to be analyzed from an internal, within-person perspective.
If a specific person changes their media diet, we need to measure how this affects _them_.
Between person comparison from cross-sectional data, where participants are interviewed only once, cannot provide such data.
In this study, I will hence differentiate between within-person effects and between-person relations.

One precondition of causality is temporal order. 
The cause needs to precede the outcome.
To this end, finding the right interval when causes and effects are measured is crucial.
For example, if we want to measure the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measure one minute, one hour, one day, or one week after consumption.
Finding the right interval is difficult.
If variables are more stable, longer intervals are needed, and if they fluctuate a lot, shorter intervals are necessary.
In the case of well-being, to analyze affect we need shorter intervals, while for life satisfaction longer ones.
Still, finding the right interval is challenging, because especially short intervals are practically hard to implement and require experience sampling measures (also known as in situ measurement or ambulant assessment).

In this study, I will adopt an intermediate perspective.
I will analyze if, when a person changes their social media diet, are there simultaneous changes in well-being?
When additionally controlling for both stable and varying covariates, we can then be more sure that the effect is indeed causal.
This approach was implemented already by several other studies and is considered one of the best practices to analyze causality.

@@

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, p-value) were preregistered on the Open Science Framework. 
The (anonymous) preregistration can be accessed here: https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3.
Because in this study I analyze data from an already existing large scale data set, the Austrian Corona Panel Project, all of these steps were preregistered prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online [@kittelAustrianCoronaPanel2020].
In some cases the analyses could not be executed as originally planned, because some properties of the variables only became apparent upon data analysis.
The most relevant deviations are reported below, and a complete list of all changes can be found online.

@@

The sample was representative of the Austrian population in terms of age, gender, region/state, municipality size, and educational level. In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years old. 
The average age was `r mean(2021 - d_wide$year_birth, na.rm = T) %>% round(0)` years, `r mean(d_wide$male, na.rm = T) %>% round(., 2) * 100` percent were male, `r d_wide %>% filter(.$edu_fac %in% c("Hochschulverwandte Lehranstalt oder Kolleg","Bachelor", "Magister / Master / Diplomingenieur / Fachhochschule", "Doktor / PhD")) %>% nrow() / sum(table(d_wide$edu_fac)) %>% divide_by(100)` percent had a University degree, and `r table(d_wide$employment_fac)["Arbeitslos"] / (table(d_wide$employment_fac) %>% sum()) %>% unname() %>% divide_by(100) %>% round(., 0)` percent were currently unemployed.
<!-- The study received IRB approval from the University of Vienna. -->

Because the data were analyzed post hoc, no sample size planning on the basis of a priori power analyses was conducted.
The sample is very large, and it is hence well-equipped reliably to detect also small effects, which is why no exact post hoc power analysis were conducted.
In addition, because such large samples easily generate significant _p_-values even for very small effects, this study builds on a smallest effect size of interest.
To test the hypotheses, I will use the interval testing approach as proposed by @dienesUsingBayesGet2014.
On the basis of the SESOI, I will define a null region.
For well-being, the null region will be between _b_ = -.30 and _b_ = .30.
For example, if the 95% confidence interval lies completely within the null-region (e.g., _b_ = .15, [95% CI: .05, .25]), the hypothesis that the effect is only trivial is be supported.
If the effects interval and the null region overlap (e.g., _b_ = .25, [95% CI: .15, .35]), the hypothesis is not supported and the results are considered inconclusive, whereas a meaningful negative effect is rejected.
If the confidence falls completely outside the null-region (e.g., _b_ = .4, [95% CI: .35, .45]), the hypothesis is rejected and the existence of a meaningful positive effect is supported.

@@

The hypothesis was analyzed using mixed effects models, namely random effect within-between models (REWB)[@bellFixedRandomEffects2019]. 
Three models were run, one for each dependent variable.
All predictors---that is, social media activities and social media channels, within and between-person factors, stably and varying covariates---were included in each of the three models. 
The data were hierarchical, and responses were nested in participants and waves.
Nesting in participants allowed to separate within-person effects from between-person relations.
Nesting in waves allowed to control for general developments, such as general decrease in well-being in the population, for example due to lockdown measures (hence, there was no need additionally to control for specific phases or measures of the lockdown).

The factorial validity of the scales were tested with confirmatory factor analyses (CFA). 
If Mardia’s test showed that the assumption of multivariate normality was violated, the more robust Satorra-Bentler scaled and mean-adjusted test statistic (MLM) was used as estimator. 
To avoid overfitting, scales were tested on more liberal fit criteria (CFI > .90, TLI > .90, RMSEA <. .10, SRMR < .10) [@klinePrinciplesPracticeStructural2016].
Because REWB-models cannot model scales as latent variables, to increase precision factor scores exported from the CFAs were used.
For more information on the analyses, see [companion website](https://xmtra.github.io/Austrian_Corona_Panel_Project/index.html).

@@

In what follows, I briefly list all variables that were collected.
For a complete list of all items and item characteristics, see [companion website](https://xmtra.github.io/Austrian_Corona_Panel_Project/index.html).

@@

All three items were measured on each wave.

### COVID-19 related social media use

The COVID-19 related social media use focused on interaction was measured with three variables (a) reading, (b) liking and sharing, and (c) posting.
The general introductory question was "How often during the last week have you engaged in the following activities on social media?"
The three items read as follows: 
"Reading the posts of others with content on the Coronavirus"; "When seeing Posts on the Coronavirus, I clicked 'like', 'share' or 'retweet'"; "I myself wrote posts on the Coronavirus on Social Media."
Answer options were 1 (_several times per day_), 2 (_daily_), 3 (_several times per week_), 4 (_weekly_), 5 (_never_).
The items were inverted for the analyses.
The variable's average across all waves was _M_ = `r get_specs(model_life_sat)$mean`, ranging from _M_~min~ = `r get_specs(model_life_sat)$min` to _M_~max~ = `r get_specs(model_life_sat)$max`.
The average standard deviation was _SD_ = `r get_specs(model_life_sat)$sd`.

The COVID-19 related social media use focused on channels was measured with five variables.
The general introductory question was "How often in the last week have you followed information related to the Corona-crisis in the following social media?"
The five items were (a) Facebook, (b) Twitter, (c) Instagram, (d) Youtube, (e) WhatsApp.
Again, the answer options were 1 (_several times per day_), 2 (_daily_), 3 (_several times per week_), 4 (_weekly_), 5 (_never_).
Again, the items were inverted for the analyses.

All items were measured every second wave.
However, freshly recruited respondents always answered these questions.

@@

The effects of COVID-19 related social media use were controlled for the following stable variables: 
(a) gender (answer options: female, male, diverse), (b) age, (c) education (10 ordinal options), (d) Austria country of birth (yes/no), (e) Austria parents' country of birth (no parent, one parent, both parents).
We controlled also for the following varying covariates:
(a) text-based media news consumption, (b) video-based media news consumption, (c) residency is Vienna, (d) household size, (e) Self-reported physical health, (f) Living space (in squaremeter), (g) access to balcony, (h) access to garden, (i) Employment status, (j) Work hours per week, (k) Home office, (l) household income, (m) outdoor activities, (o) satisfaction with democracy, (p) disposition to take risks, (q) locus of control.

@@

When looking at the variables from a descriptive perspective, we see that all well-being measures did not change substantially across the different phases of the pandemic.
COVID-19 related media use, however, increased during the beginning of the study and remained stable after approximately six waves.

@@

The study's hypothesis was that the effects of all types of social media use on well-being will be trivial.
Regarding the different types of _communication_, that is reading vs. sharing vs. posting, all within-person effects fell completely within the a-priori defined SESOIs.
For example, respondents who used social media more frequently than usual to read about COVID-19 related topics did not show a simultaneous change in life satisfaction (_b_ = `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Reading" & type == "within") %>% select(conf.high) %>% round(2)`]).
Only one effect did not include zero.
If people posted more about COVID-19 related aspects than they usually did, life satisfaction increased (_b_ = `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Life satisfaction" & iv == "Posting" & type == "within") %>% select(conf.high) %>% round(2)`]).
However, because the effect did not fall outside our null region, it's likely not large enough to be considered meaningful.
As a result, the hypothesis was supported for all types of COVID-19 related social media communication.

Regarding between-person relations, about which no hypotheses were formulated, only two effects didn't include zero.
First, respondents who across all waves used social media more frequently than others to read about COVID-19 reported higher levels of positive affect than others (_b_ = `r dat_fig_results_activity %>% filter(dv == "Positive affect" & iv == "Reading" & type == "between") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Positive affect" & iv == "Reading" & type == "between") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Positive affect" & iv == "Reading" & type == "between") %>% select(conf.high) %>% round(2)`]).
However, note that this effect still fell completely within the null-region.
Hence, although positive the effect was considered too small to matter practically.
Second, respondents who across all waves used social media more frequently than others to read about COVID-19 reported lower levels of negative affect than others (_b_ = `r dat_fig_results_activity %>% filter(dv == "Negative affect" & iv == "Reading" & type == "between") %>% select(estimate) %>% round(2)` [95% CI `r dat_fig_results_activity %>% filter(dv == "Negative affect" & iv == "Reading" & type == "between") %>% select(conf.low) %>% round(2)`, `r dat_fig_results_activity %>% filter(dv == "Negative affect" & iv == "Reading" & type == "between") %>% select(conf.high) %>% round(2)`]).
The effect was partially outside the null region, hence the effect might be large enough to be practically relevant, and the effect that it's trivial cannot be rejected.

@@

Using a panel study with a representative sample of the Austrian population that consistent of 24 waves, this study analyzed the effects of COVID-19 related social media use on well-being.
A random effect model that separated between person relations from within-person effects and that controlled for several third variables showed that within-person effects were trivial.
People who used social media more than usual to learn about COVID-19 did not show changes in their well-being.
As a result, the results imply that COVID-19 related social media use is irrelevant for people's well-being.
Other factors among the third variables that were measured revealed much larger effects, implying that the relevance of specific types of social media use for well-being is limited.
Popular fears that "doomscrolling" or overusing social media during times of crises might not be justified.

The study is not aligned with a recent cross-sectional study by @bendauAssociationsCOVID19Related2021 that showed negative relations between social media and well-being.
However, @bendauAssociationsCOVID19Related2021 analyzed cross-sectional data on a between-person level, while not controlling for third variables, which does not allow to make causal inferences.
At the same time, this study is well-aligned with recent studies and meta-analyses from related research questions, which found that the effects of various types of social media use on several well-being indicators is small at best, often too small to matter [@meierComputerMediatedCommunicationSocial2020; @orbenTeenagersScreensSocial2020; @fergusonThisMetaanalysisScreen2021].
